---
title: "Running a geneclade simulation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{geneclade_vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Geneclade is a simulation-based model that seamlessly integrates eco-evolutionary processes taking place at both population and species levels. Our simulation runs in continuous time, which allows us to model processes of very different temporal scales together within a single framework. While tracking each individual for multiple populations simultaneously for multiple species over time is computationally infeasible, we tracked dynamics within populations that are the product of the behavior of individuals. At each cycle in the simulation, a species is selected to undergo an event (i.e., either a population or species-level process). The probability of each species being selected depends on its number of populations, and the frequency of events depends on the rates used. Rates in our model are the number of events taking place per population per time unit.

The main function is run_geneclade_simulation() and the documentation for that function (using ?run_geneclade_simulation) should be enough to run it. I will use this vignette to further expand the explanation for some arguments.

Geneclade offers different ways to stop the simulation: based on time, reaching a dynamic equilibrium or a given number of species. We need to set the argument "max_spp" which indicates the desired number of species to have in the simulation as long as "condition_to_stop" is set to "richness" . Once the number of species is reached, no further speciation events will take place. Then, simulation will carry on until a dynamic equilibrium is met. However, if we set condition_to_stop = "time", we will have as many speciation events as expected given 1) the "rate_speciation", and 2) the "simulated_time." Once the simulated time has been reached, no further speciation will take place and the simulation will run until the dynamic equilibrium is reached. In this case, the time in the simulation might go way beyond "simulated_time". If one wants to force the simulation to stop once certain time is reached (the simulated_time) regardless of finding or not an equilibrium, use stop_time_after_change = 0.01.

```{r setup}
library(geneclade)
max_spp <- 20
condition_to_stop <- "richness"
stop_time_after_change <- 0
```

We can ask the software to store what the simulation is like at specific time slices for further analysis of changes in the simulated world over time. The output at time slices and at the end of the simulation will be have the same structure: 1) a table with geographic and population genetic information, 2) a phylogenetic tree to the clade radiation, and 3) a table to use in ggplot to make diversity and abundance maps. The time slices are specified with "time_slices". Notice that it is difficult to estimate what the total simulated time would be to achieve equilibrium. This means that, for instance, when simulation will run 20000 time units to equilibrium, the argument time_slices <- c(10,20,30) will store 3 time slices at the very beginning of the simulation with nothing to store later in the simulation. Remember that when simulation stops, the end state of the system will always be stored, even if time_slices <- 0. We recommend making a sensible decision on the number of time slices to request, as there must be a balance between how informative the temporal analysis is and the computing power needed to process all the time slices.

```{r}
time_slices <- c(10,200,600,1000)
```

The rest of arguments to the function run_geneclade_simulation are rather straightforward.

```{r}
position_start_x <- 17
position_start_y <- 17
simulated_time <- 2500
rate_speciation <- 0.00005
rate_colonisation <- 5
rate_extirpation <- 0.0000001
rate_geneflow <- 0.01
rate_demographicchange <- 1
rate_traitevolution <- 0
rate_mutation <- rate_geneflow/10
alleles_optimum_enviroment <-  rep(9,25)
vicariant_speciation <- TRUE
maximum_cycles <- 200000
frequency_print_simulationstatus <- 10000
```

For the map, one needs to specify the dimensions in x and y, i.e., the number of cells in x and the number of cells in y. The arguments "x_max" and "y_max" should match the maps. Notice that I have pre-loaded maps for this vignette so I am using the function get().
```{r}
x_max <- 42
y_max <- 42
map_k_1 <- get("map_k_1")
# one normally loads the map with:
#map_k_1 <- read.table(paste0("k_map_uniform.txt"))
map_environment_1 <- get("map_environment_1")
#map_environment_1 <- read.table(paste0("temperature_map_uniform.txt"))
```

Finally, we call the function:
```{r}
output_simulation <- run_geneclade_simulation (position_start_x,
                                     position_start_y,
                                     advanced_initialization = NULL,
                                     max_spp,
                                     simulated_time,
                                     condition_to_stop ,
                                     stop_time_after_change,
                                     time_slices,
                                     maximum_cycles ,
                                     rate_speciation,
                                     rate_colonisation,
                                     rate_extirpation,
                                     rate_geneflow,
                                     rate_demographicchange,
                                     rate_traitevolution,
                                     rate_mutation,
                                     alleles_optimum_enviroment,
                                     percentage_geneflow = 10,
                                     vicariant_speciation,
                                     manual_speciation_events_timing = 0,
                                     growth_only = TRUE,
                                     unlink_range_to = NULL,
                                     x_max,
                                     y_max,
                                     map_k_1,
                                     map_environment_1,
                                     frequency_print_simulationstatus)


```

Simulation stopped at cycle 200000 as requested, which allows to complete a simulation for this vignette in reasonable real-life time. No equilibrium was achieved in 200000 cycles. The object output_simulation contains three outputs as described above. We can see the object number three:
```{r}
phylotrees_overtime <- output_simulation$output_phylotree_overtime
phylotrees_overtime
```
This object is also a list, and it will have as many elements as time slices we requested. We can see that the first time slices were so early in the simulation, that we did not have enough species to build a tree. 
The next output object I would like to call now is the species table. Similar to above, it is a list with as many elements (tables) as time slices where requested.
```{r}
tablespecies_overtime <- output_simulation$output_table_overtime
```
We can see what the system at the end of the simulation was like by retrieving the last element of that object.  The name of each element indicates time of that slice (or final time of the simulation):

```{r}
this_time_slice <- tablespecies_overtime[length(tablespecies_overtime)]
names(this_time_slice)
```
Let us now see what the table is like. Here I show some sampled lines of it. The Sp_ID column shows the id of the species. A species will have as many rows as populations it has. So, a given row will indicate what species it belongs to, its xy location, number of individuals and the alleles that are carried. Use this table to compute population genetic metrics.
```{r}
this_time_slice <- tablespecies_overtime[[length(tablespecies_overtime)]]
sampled_lines <- sample(1:nrow(this_time_slice),10,replace = F)
sampled_lines <- sort(sampled_lines)
this_time_slice[sampled_lines,]
```
Geneclade offers a function to compute basic metrics. Now imagine we want to compute calculate Fst for the species with most populations (most widespread species). We can do:

```{r}

most_widespread <- names(sort(table(this_time_slice$Sp_ID),decreasing = TRUE)[1])
this_species <- this_time_slice[which(this_time_slice$Sp_ID == as.numeric(most_widespread)),]
#Total number of populations for this species:
nrow(this_species)
# remove columns with uneccesary info (xy coordinates for instance)
this_species_pop_info <- this_species[,-c(1,2,3,4)]
computed_genenetic_metrics <- compute_PopulationsSpecificFST(this_species_pop_info,
                                                            pairwise = FALSE,
                                                            locus="both")
# Hudson's Fst
mean(as.numeric(computed_genenetic_metrics$computed_betas$betaiovl),na.rm=TRUE)
# Weir and Cockrham Fst
computed_genenetic_metrics$wc_fst$FST
```
